Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
fix breakage of annotations
fix breakage

\begin{figure}[h]
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional}} 
%		\label{traverse_lock_a}	
	\end{subfigure}\qquad
	\begin{subfigure}[t]{0.48\textwidth}
	\begin{lstlisting}[language=caml, style=myStyle]
	let insert r k =
	  lockNode r;
	  let n = traverse r k in
	  let res = insertOp n k in
	  unlockNode n; res
	\end{lstlisting}
%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
%		\label{traverse_lock_b}
	\end{subfigure}
	\caption{An example search structure template (lock coupling)}
	\label{template-ex}
\end{figure}

@@

%maybe inline this instead of making it a figure?
\begin{figure}[ht]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
			\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
			\begin{array}{c} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
				\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the lock-coupling template}
	\label{fig:traverse_lock}
\end{figure}

@@

\begin{figure}[!ht]
	$\color{blue}
	\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
	\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
	\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
	\begin{array}{c} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
		\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
	\end{array}
	\ \Bigg| \ \treerep\ m \
	\right\rangle$
	\caption{The \texttt{traverse} function in lock-coupling template annotated with separation logic specification}
	\label{proof_lock_traverse}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  m. \left\langle
			\begin{array}{c}
				\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ \infp (\texttt{n}) \ \big| \ \treerep\ m
			\end{array}
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \ \texttt{n'}.
			\begin{array}{c}
				\texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ 
				\\ \infp(\texttt{n'}) \ \ast \ \mathsf{R}(\texttt{n'}) \ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the give-up template}
	\label{fig:traverse_giveup}
\end{figure}

@@

Figure \ref{proof_lock_insert} in Appendix \ref{sec:apd_proof} depicts the \texttt{insert} function within the lock-coupling template for BST, annotated using separation logic.
The proof for \texttt{insert} begins with the precondition $\nodeboxrep \ \texttt{p}$. We set \texttt{pn->n} to point to the root node \texttt{*r} (line 3 in Figure \ref{proof_lock_insert}). After the lock for the root node is acquired (line 5 in Figure \ref{proof_lock_insert}), the thread acquires the contents of the root node, $\mathsf{R}(\texttt{r})$. We can then satisfy the precondition of the lock-coupling \lstinline{traverse} function from Section \ref{traverse_proof_lock} \wm{Maybe we should show the spec again here}, allowing us to call \lstinline{traverse} and obtain its postcondition.
\begin{mathpar}
	% spec of inRange
	{\color{blue}
		\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
		\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
		\right\rangle
	}
	\\ 
	\texttt{traverse(pn, k)} 
	\\
	{\color{blue}
		\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
		\begin{array}{c} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
			\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
		\end{array}
		\ \Bigg| \ \treerep\ m \
		\right\rangle
	}
\end{mathpar}
In the $\mathsf{then}$ branch, when \texttt{traverse} finds a node with key \lstinline{k}, %resulting in $\texttt{res} = \texttt{false}$, 
we know that $\texttt{pn->n'->t} \neq \texttt{NULL}$, i.e., the key to be inserted already exists in the tree. In this scenario, we update the node's value while maintaining the rest of the tree structure (line 9 in Figure \ref{proof_lock_insert}). Otherwise, when \texttt{traverse} reaches a leaf and produces $\mathit{res} = \texttt{true}$, we know that $\texttt{pn->n'->t} = \texttt{NULL}$ (at line 14-15 in Figure \ref{proof_lock_insert}). We can then call the \texttt{insertOp} specification, adding the new key-value pair to the tree. In either case, before we release the node's lock, we must show that we have altered the tree from its current abstract state $m$ to $m[\texttt{k} \mapsto \texttt{v}]$, thereby satisfying the public postcondition of \lstinline{insert}. We do this by demonstrating that the in-memory tree update corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the tree. 
We can then release the lock (line 21 in Figure \ref{proof_lock_insert}) return the node's contents to the shared state. %Subsequently, we demonstrate the postcondition of the \texttt{insert} function to complete the proof.
%\wm{Do we ever define $\treerep$?}

@@

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% inRange for give-up
		{\color{blue}
			\left\{ 
			\begin{array}{c}
				\texttt{pn->n}\mapsto\texttt{t} \ \ast \ \texttt{pn->n}\mapsto\texttt{min} \ \ast \  \texttt{pn->n}\mapsto\texttt{max} 
			\end{array}
			\right\}
		}
		\\ 
		\texttt{inRange(pn, k)} 
		\\
		{\color{blue}
			\left\{\mathit{res.} \
			\begin{array}{c}
				\texttt{pn->n}\mapsto\texttt{t} \ \ast \ \texttt{pn->n}\mapsto\texttt{min} \ \ast \  \texttt{pn->n}\mapsto\texttt{max} \  \ast  \ 
				\\(\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (\texttt{min} < \texttt{k} < \texttt{max})\ \mathsf{else}\ (\texttt{k} \leq \texttt{min} \lor \texttt{k} \geq \texttt{max}))
			\end{array}
			\right\}
		}
	\end{mathpar}
	\caption{Specifications of \texttt{inRange} for the give-up template}
	\label{fig:inRange_giveup}
\end{figure}

@@

\begin{figure}[!ht]
	$\color{blue}
	\forall \  m. \left\langle
	\begin{array}{c}
		\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ \infp (\texttt{n}) \ \big| \ \treerep\ m
	\end{array}
	\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \ \texttt{n'}.
	\begin{array}{c}
		\texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ 
		\\ \infp(\texttt{n'}) \ \ast \ \mathsf{R}(\texttt{n'}) \ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
	\end{array}
	\ \Bigg| \ \treerep\ m \
	\right\rangle$
	\caption{The \texttt{traverse} function in give-up template annotated with separation logic specification}
	\label{proof_giveup_traverse}
\end{figure}

>>>>

\begin{figure}[h]
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional}} 
%		\label{traverse_lock_a}	
	\end{subfigure}\qquad
	\begin{subfigure}[t]{0.48\textwidth}
	\begin{lstlisting}[language=caml, style=myStyle]
	let insert r k =
	  lockNode r;
	  let n = traverse r k in
	  let res = insertOp n k in
	  unlockNode n; 
	  res\end{lstlisting}
%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
%		\label{traverse_lock_b}
	\end{subfigure}
	\caption{An example search structure template (lock coupling)}
	\label{template-ex}
\end{figure}

@@

%maybe inline this instead of making it a figure?
\begin{figure}[ht]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
			\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
			\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
				\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the lock-coupling template}
	\label{fig:traverse_lock}
\end{figure}

@@

\begin{figure}[!ht]
	$\color{blue}
	\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
	\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
	\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
	\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
		\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
	\end{array}
	\ \Bigg| \ \treerep\ m \
	\right\rangle$
	\caption{The \texttt{traverse} function in lock-coupling template annotated with separation logic specification}
	\label{proof_lock_traverse}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  m. \left\langle
			\begin{array}{c}
				\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ \infp (\texttt{n}) \ \big| \ \treerep\ m
			\end{array}
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \ \texttt{n'}.
			\begin{array}{l}
				\texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ 
				\\ \infp(\texttt{n'}) \ \ast \ \mathsf{R}(\texttt{n'}) \ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the give-up template}
	\label{fig:traverse_giveup}
\end{figure}

@@

Figure \ref{proof_lock_insert} in Appendix \ref{sec:apd_proof} depicts the \texttt{insert} function within the lock-coupling template for BST, annotated using separation logic.
The proof for \texttt{insert} begins with the precondition $\nodeboxrep \ \texttt{p}$. We set \texttt{pn->n} to point to the root node \texttt{*r} (line 3 in Figure \ref{proof_lock_insert}). After the lock for the root node is acquired (line 5 in Figure \ref{proof_lock_insert}), the thread acquires the contents of the root node, $\mathsf{R}(\texttt{r})$. We can then satisfy the precondition of the lock-coupling \lstinline{traverse} function from Section \ref{traverse_proof_lock} \wm{Maybe we should show the spec again here}, allowing us to call \lstinline{traverse} and obtain its postcondition.
\begin{mathpar}
	% spec of inRange
	{\color{blue}
		\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
		\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
		\right\rangle
	}
	\\ 
	\texttt{traverse(pn, k)} 
	\\
	{\color{blue}
		\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
		\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
			\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
		\end{array}
		\ \Bigg| \ \treerep\ m \
		\right\rangle
	}
\end{mathpar}
In the $\mathsf{then}$ branch, when \texttt{traverse} finds a node with key \lstinline{k}, %resulting in $\texttt{res} = \texttt{false}$, 
we know that $\texttt{pn->n'->t} \neq \texttt{NULL}$, i.e., the key to be inserted already exists in the tree. In this scenario, we update the node's value while maintaining the rest of the tree structure (line 9 in Figure \ref{proof_lock_insert}). Otherwise, when \texttt{traverse} reaches a leaf and produces $\mathit{res} = \texttt{true}$, we know that $\texttt{pn->n'->t} = \texttt{NULL}$ (at line 14-15 in Figure \ref{proof_lock_insert}). We can then call the \texttt{insertOp} specification, adding the new key-value pair to the tree. In either case, before we release the node's lock, we must show that we have altered the tree from its current abstract state $m$ to $m[\texttt{k} \mapsto \texttt{v}]$, thereby satisfying the public postcondition of \lstinline{insert}. We do this by demonstrating that the in-memory tree update corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the tree. 
We can then release the lock (line 21 in Figure \ref{proof_lock_insert}) return the node's contents to the shared state. %Subsequently, we demonstrate the postcondition of the \texttt{insert} function to complete the proof.
%\wm{Do we ever define $\treerep$?}

@@

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% inRange for give-up
		{\color{blue}
			\left\{ 
			\begin{array}{c}
				\texttt{pn->n}\mapsto\texttt{t} \ \ast \ \texttt{pn->n}\mapsto\texttt{min} \ \ast \  \texttt{pn->n}\mapsto\texttt{max} 
			\end{array}
			\right\}
		}
		\\ 
		\texttt{inRange(pn, k)} 
		\\
		{\color{blue}
			\left\{\mathit{res.} \
			\begin{array}{l}
				\texttt{pn->n}\mapsto\texttt{t} \ \ast \ \texttt{pn->n}\mapsto\texttt{min} \ \ast \  \texttt{pn->n}\mapsto\texttt{max} \  \ast  \ 
				\\(\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (\texttt{min} < \texttt{k} < \texttt{max})\ \mathsf{else}\ (\texttt{k} \leq \texttt{min} \lor \texttt{k} \geq \texttt{max}))
			\end{array}
			\right\}
		}
	\end{mathpar}
	\caption{Specifications of \texttt{inRange} for the give-up template}
	\label{fig:inRange_giveup}
\end{figure}

@@

\begin{figure}[!ht]
	$\color{blue}
	\forall \  m. \left\langle
	\begin{array}{c}
		\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ \infp (\texttt{n}) \ \big| \ \treerep\ m
	\end{array}
	\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \ \texttt{n'}.
	\begin{array}{l}
		\texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ 
		\\ \infp(\texttt{n'}) \ \ast \ \mathsf{R}(\texttt{n'}) \ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
	\end{array}
	\ \Bigg| \ \treerep\ m \
	\right\rangle$
	\caption{The \texttt{traverse} function in give-up template annotated with separation logic specification}
	\label{proof_giveup_traverse}
\end{figure}
